An item response theory analysis of the matrix reasoning item bank (MaRs-IB)

### 2022-01-18 manuscript/main.tex
more edits

Across all 384 unique item variants, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74) (Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) had accuracy rates beneath chance levels (25\%), and only 22 items (5.7\%) had performance levels above 90\%. Mean performance on an item was significantly and negatively correlated with the number of changing relationships amongst its features (Spearman's $\rho = -0.482, p < 1e^{-6})$). Moreover, item-wise accuracy rates in this sample were remarkably similar to those reported by \cite{chierchia2019matrix} (Spearman's $\rho = 0.870, p < 1e^{-6})$), when restricting the latter to items with more than 200 available responses. These results further corroborate \cite{chierchia2019matrix} in that the majority of items exhibited good performance, with few items at ceiling or floor and passing of the manipulation check.

@@

Models were compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the conditional leave-one-cluster out (LOCO) cross- validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses).

@@

First we estimated Models 1-4 holding constant the item discrimination parameters ($\alpha = 1$), which allowed us to investigate the difficulty of MaRs-IB items independently. (Note that these models still included guessing parameters fixed to the nominal guessing rate. We also estimated models with no guessing parameter, i.e. 1PL or Rasch models, but model comparison revealed these to be inferior.)

The goodness of fit of the four models was compared based on leave-one-cluster out (LOCO) cross-validation as summarized in Table \ref{table:1}. The winning model was Model 1d, in which item difficulties were modeled as a function of all level-1 and level-2 fixed and random effects. This has multiple interpretations, which we will unpack in turn. First, this implies that there was systematic variability in item difficulty at the family-level beyond what could explained by their attributes (this is reflected...

[above to be combined with below when time comes]

First, Model 2 was preferred to Model 1 suggesting that there is variability in the difficulty of item families above and beyond what can be explained by their attributes. Similarly, Model 3 was preferred to Model 2 suggesting that there is systematic variability in the clones of each item family. Finally, Model 4 was preferred to Model 3 suggesting that there is variability in the difficulty of item clones above and beyond what can be explained by their respective attributes. In summary, the full model -- including all fixed and random effect terms -- was preferred to the three simpler models, indicating that item families and clones are nonequivalent even after controlling for their attributes.

The average item difficulty was $\beta_\mu = 0.177$ (95\% HDI = 0.118 - 0.238) but this ranged considerably across items (sd = 1.431, 95\% HDI = 1.370 - 1.492).

Over half of the variability in item difficulty (65.6\%) was explained by the fixed effects of item family (level-1) and clone (level-2) attributes. The associations between level-1 attributes and item difficulty were as expected: feature number was positively associated with difficulty (beta), such that the inclusion of one additional feature was associated with a 13.1\% reduction in expected accuracy. Similarly, rule number was positively associated with item difficulty (beta), such that the inclusion of one additional rule a 10.1\% reduction in expected accuracy. Turning next to level-2 attributes, distractor type was unexpectedly associated with item difficulty (beta), such that clones with MD distractors clones were associated with a 13.6\% reduction in accuracy compared to clones with PD distractors. This pattern is clearly evident in the raw item-level scores (Figure XX). Finally, average response time was positively associated with difficulty, such that a one standard deviation change in response time was associated with an 11.8\% reduction in accuracy.

The inclusion of item family (level-1) random effects explained an additional 18.5\% of the variance in item difficulty. In total, the percentage of variance in item difficulty not explained by residual variance from item clones (level-2) was 84.1\%. The residual variance was harder to explain (Figure XXb); indeed, there was no clear patterns in difficulty by item set, suggesting that at least part of residual variance may be explained by idiosynchratic perceptual features of item clones.

Next we estimated three models as above, but with two key differences: First, we no longer held constant the item discrimination parameters but rather estimated them alongside the item difficulty parameters. Second, whereas the item difficulty parameters were always estimated using the full additive multilevel item structure model (i.e. level-1/2 fixed and random effects), the item discrimination parameters were estimated adding one additional component per model (Model 1: level-1/2 fixed effects only; Model 2: fixed effects and level-1 random effects; Model 3: full model). The goodness of fit of the three models was again compared based on leave-one-cluster out (LOCO) cross-validation as summarized in Table \ref{table:2}.

First, all three models including discrimination parameters were preferred to the best-fitting model with discrimination parameters held constant. Model 2 was preferred to Model 1 suggesting that there is variability in the discrimination of item families not accounted for by their attributes. Importantly, Model 2 was preferred to Model 3 such that there was insufficient variability in discrimination in item clones to necessitate additional parameters. We return to this point in the Discussion.

Returning to variance partitioning, just less than half of the variance in item discrimination (44.9\%) was explained by attributes of the explained by attributes of item families and clones. Interestingly, only the number of item rules was positively associated with item discrimination (HDI); no other attributes were credibly related (HDIs here).

Finally, we turn to explanatory item response modeling. The largest effects were from the response times. Average response time blah. Delta response time blah. Interestingly, we found a negative correlation between age and ability, such that older participants tended to perform worse on average. Finally, we observed a small but credible association between gender and ability. Though we used the colorblind stimulus set, it is possible there was residual perceptual difficulty for men compared to women. Granted, this effect is small.

This paragraph should be on fit. Should mention: simulation study, posterior predictive check, item fit indices, and unidimensionality measure. You should probably review a couple of other RPM validation papers to make sure our methods are straight (e.g. doi.org/10.1037/a0027830)

@@

In discussing the success of the MaRs-IB generation process, some things to point out:

1. A lot of variance in item difficulty and discrimination can be explained by inherent features of the items. This is great!

2. Though there is quite a bit of residual variance at the item family level, this is fine as we only modeled two features. A point we return to in the Discussion.

3. At the clone level, it seems clear that item clones are not equivalent w.r.t. difficulty. For one thing, MD vs. PD big. Should describe level-2 standard deviation somewhere.

4. For clone discrimination, it seems less clear. One possibility is that there isn't much variability (cite Embertson). Another possibility is that we are underpowered. Cite Lathrop here for discussion of potential bias.

Summarize as good to go!

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrrrrr}
\hline
&  Coef. & Std.Err. & z & P$> |$z$|$ & [0.025 & 0.975]  \\
\hline
Intercept &  2.648 &    0.009 & 287.757 &       0.000 &  2.630 &  2.666  \\
accuracy &  0.039 &    0.007 &   5.367 &       0.000 &  0.025 &  0.053  \\
difficulty &  0.073 &    0.005 &  13.292 &       0.000 &  0.062 &  0.083  \\
accuracy:difficulty &  0.097 &    0.007 &  14.351 &       0.000 &  0.084 &  0.110  \\
score &  0.167 &    0.009 &  19.204 &       0.000 &  0.150 &  0.184  \\
accuracy:score & -0.064 &    0.007 &  -8.764 &       0.000 & -0.078 & -0.050  \\
difficulty:score &  0.032 &    0.005 &   6.549 &       0.000 &  0.023 &  0.042  \\
accuracy:difficulty:score &  0.012 &    0.006 &   1.955 &       0.051 & -0.000 &  0.025  \\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

Across all 384 unique item variants, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74) (Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) had accuracy rates beneath chance levels, and only 22 items (5.7\%) had performance levels above 90\%. Mean performance on an item was significantly and negatively correlated with the number of changing relationships amongst its features (Spearman's $\rho = -0.482, p < 1e^{-6})$). Moreover, item-wise accuracy rates in this sample were remarkably similar to those reported by \cite{chierchia2019matrix} (Spearman's $\rho = 0.870, p < 1e^{-6})$), when restricting the latter to items with more than 200 available responses. These results further corroborate \cite{chierchia2019matrix} in that the majority of items exhibited good performance, with few items at ceiling or floor and passing of the manipulation check.

@@

The goodness of fit of the models was compared were compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. Importantly, we computed the conditional leave-one-cluster out (LOCO) cross- validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses).

@@

First we estimated the four models described above holding the item discrimination parameters fixed ($\alpha = 1$).  Note that these models still included a fixed guessing parameter ($\gamma = 0.25$). We also estimated models with no guessing parameters (i.e. one-parameter logistic or Rasch models), but model comparison indicated these models were worse fits to the data (Table XX).

The goodness of fit of the four models was compared using Bayesian leave-one-cluster out cross-validation as summarized in Table \ref{table:1}. The full model (level-1/2 fixed \& random effects) fit the response data better than the model with only level-1 fixed effects ($\Delta$ LOCO = 1223.69, se = 44.94), the model with level-1 fixed and random effects ($\Delta$ LOCO 511.58, se = 28.77), and the model with all but level-2 random effects ($\Delta$ LOCO = 187.96, se = 17.25). Therefore, the two-level model with fixed and random item difficulty effects is chosen as the estimated model to discuss further.

Overall, the average item difficulty was $\mu_\beta = 0.177$ (95\% HDI = 0.118 - 0.238). Difficulty also ranged considerably across items (sd = 1.431, 95\% HDI = 1.370 - 1.492). We next unpack the sources of variability in item difficulty, beginning first with the level-1 attributes. As expected, a one-unit change in the number of item features was credibly associated with an increase in item difficulty ($\delta_1$ = 0.132, 95\% HDI = 0.087 - 0.174). Similarly, a one-unit change in the number of changing relations was also associated with an increase in item difficulty ($\delta_2$ = 0.112, 95\% HDI = 0.080 - 0.144). Together, these two attributes explained approximately 65.5\% of the variance in difficulty across item families (35.0\% across all items). This result supports the success of the item generation process, insofar that item difficulty is associated with and well-explained by the core item attributes. It also highlights, however, that item families with identical attributes are not exchangeable as confirmed by the better fit of Model 1b compared to Model 1a ($\Delta$ LOCO = 712.11, se = 34.61).

Turning next to the level-2 attributes, there was an unexpected association between distractor type and item difficulty such that otherwise identical items were more difficult with MD than PD distractors ($\delta_3$ = 0.252 , 95\% HDI = 0.218 - 0.287). There was also an association between average response time and item difficulty ($\delta_4$ = 0.122, 95\% HDI = 0.093 - 0.151). In total, the controllable fixed effects (number of features, number of changing relations, distractor type), explained nearly half (48.2\%) of the variance in item difficulty. Compared to a model with only level-1 fixed and random effects, the improved fits of Models 1c ($\Delta$ = 323.62, se = 23.36) and 1d ($\Delta$ LOCO = 187.96, se = 17.25) indicates that item clones are not equivalent in their difficulty. Taking into account the controllable fixed effects and level-1 random effects, the total explained variance in item difficulty is approximately 72.0\%. To summarize, although item variants are not strictly exchangeable, only the minority of variance in item difficulty can be attributed to unmodeled effects (e.g. perceptual features).

Next we estimated the item cloning models again, but no longer holding the item discrimination parameters constant. In all models, the item difficulty parameters were estimated with fixed and random effects at both levels. In contrast, for Model 2a the item discrimination parameters were modeled as a function of level-1 and level-2 fixed effects only. In Model 2b, the item discrimination parameters were modeled as a function of the fixed effects and level-1 random effects. Finally, in Model 2c, the item discrimination parameters were modeled with fixed and random effects at both levels.

The goodness of fit of the three models is summarized in Table \ref{table:1}. Notably, the best-fitting model was not the full model but instead the model with only level-1 random effects (Model 2b). This was preferred to the model with only fixed effects ($\Delta$ LOCO = 2.13, se = 2.04) and the model with both level-1 and level-2 random effects ($\Delta$ LOCO = -1.44, se = 0.85). We note that the authors of the LOO metric recommend that standard errors be at least twice as large as the difference between models to be confident that one model truly provides a better fit. \footnote{\url{https://discourse.mc-stan.org/t/loo-comparison-in-reference-to-standard-error/4009}} As such, we proceed by interpreting Model 2b as the overall best-fitting model with caution. Further note that this model still fit substantially better than the winning model with fixed item discrimination parameters ($\Delta$ LOCO = 85.91, se = 6.33).

Overall, the average item discrimination was $\mu_\alpha = 1.298$ (95\% HDI = 1.212 - 1.385), and exhibited modest variance across items (sd = 0.221, 95\% HDI = 0.121 - 0.322). Among the fixed-effects, there was only a credible association between the number of changing relations and item discrimination ($\delta_{\alpha_2}$ = 0.067, 95\% HDI = 0.011 - 0.123). There was no such association for the number of item features ($\delta_{\alpha_1}$ =-0.037 95\% HDI = -0.090 - 0.017), distractor type ($\delta_{\alpha_3}$ -0.036, 95\% HDI = -0.082, 0.006), or average response time ($\delta_{\alpha_4}$ = -0.025, 95\% HDI = -0.075 - 0.024). Regardless, the controllable fixed effects (number of features, number of changing relations, distractor type) explained approximately (39.2\%) of the variance in item discrimination. In summary then, there is evidence that items vary in their discrimination but not strong evidence to suggest that item variants vary substantially from one another. We return to this point in the discussion.

Briefly, we turn to the person-level predictors of performance. Corroborating the mixed-effects regression model, average response time was positively associated with ability ($\rho_1$ = 0.427, 95\% HDI = 0.374 - 0.478), as was $\Delta$ response time ($\rho_2$ = 0.368, 95\% HDI = 0.313 - 0.424). There was a negative relationship between age and ability ($\rho_3$ = -0.299, 95\% HDI = -0.353 -- -0.246). Finally, there was a small and unexpected association between gender and ability ($\rho_4$ = 0.128, 95\% HDI = 0.078 - 0.182), such that women performed slightly better than men on average. Though we used the colorblind-friendly MaRs-IB stimuli for this study, it is possible that residual effects may persist.

Finally, we validated the fit of the winning model using several posterior predictive model checking measures. Across the 384 items, the average posterior predictive p-value for the $S-\chi^2$ index of item fit was $p = 0.510$ (sd = 0.242). Only 14 items (3.64\%) indicated misfit with posterior predictive p-values less than 0.05 or greater than 0.95. Moreover, the posterior predictive p-value for the test of undimensionality was $p = 0.803$. Though this suggests some evidence for local dependence, it is within a tolerable range. Overall then, the best-fitting model provides an adequate fit to the data.

@@

In this first study, we recruited a large sample of adults to complete items from the MaRs-IB, making sure that each item (and its variants) were completed by a sizable number of participants. Using hierarchical item response models, we then measured the psychometric properties (i.e. difficulty and discrimination) of each item. Corroborating the initial study from \cite{chierchia2019matrix}, we found that the items of the MaRs-IB vary greatly in their difficulty -- a prerequisite for measuring nonverbal reasoning across the ability spectrum. We also found that the items of the MaRs-IB exhibited medium-to-large levels of discrimination \citep{baker2017basics}, similar to what has been reported in previous item response theory analyses of the Raven's progressive matrices \citep{chiesi2012using}.

Moreover, using item cloning models we were able to test several important properties of the MaRs-IB items. First, we confirmed that item difficulty is a function of and to a moderate extent explained by its attributes (namely, the number of features and the number of changing relations). Relatedly, we identified that an item's discrimination is a function and partially explained by its number of changing relations, which is similar to what has been reported in other investigations of matrix reasoning tasks \citep{embretson1999generating}. Second, we found that item variants are not equivalent and exchangeable with respect to difficulty. Unexpectedly, and contrary to \cite{chierchia2019matrix}, distractor type emerged as a strong predictor of item difficulty. All else being equal, item variants using MD distractors were harder than the same items using PD distractors. Importantly, once taking this (and the residual variance in difficulty at the item family level) into account, a large majority of item difficulty (72.0\%) was accounted for, suggesting that a minority of variance in difficulty across item variants stems from unmodeled features such as their superficial perceptual properties.

summary sentence here saying that the MaRs-IB has many desirable properties. We next move onto test construction.

@@
